A North Sámi to South Sámi machine translation prototype

### 2014-02-09 paper/paper.tex
Litt mer

\subsubsection{Lexical transfer}
%@lene
Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
ex: nob `å trekke lodd' sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
fremdeles må forklare innholdet.

The bilingual dictionary
general dictionary
general words
special domain

@@

\textbf{Problems combining sma-nob and sme-nob dictionary}
This was done, it had to be manually edited because of nob-words with more than one meaning. 
Example: nob \textit{regne} means both 'to rain' and ‘to calculate’, 
which are two different verbs both in sme \textit{arvit, rehkenastit} and in sma \textit{abrodh, ryöknedh}

>>>>

\subsubsection{Lexical transfer}
%@lene

Some numbers -- kladd
\begin{itemize}
\item 3568 general word pairs + 61 proper nouns
\item added: 873 special domain word pairs 
\item added: 10 721 proper nouns
\end{itemize}

There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a North Saami-Norwegian dictionary and a South Saami-Norwegian dictionary. The word pairs were manually edited and the work revealed many incorrect pairs because of nob-words with more than one meaning. Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}. The result of this work was 3568 general word pairs and 61 proper nouns.

Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
ex: nob `å trekke lodd' sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma fremdeles må forklare innholdet ved hjelp av flere ord.

The bilingual dictionary \\
general dictionary\\
general words\\
special domain

@@
